A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-18 tex/codatadp.tex
compile
Add model completion example, more work on proof.

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not bisimilar to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.
We first state the following properties of $\Val^\ast$, 
where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

To show 2, 
first note that since \rn{Clash} does not apply to $\Ec$,
for all equivalence classes $\ec{y}$,
each pair of constructor terms $\const{C}_i( \vec t )$ and $\const{C}_j( \vec u )$ in $\ec{y}$
are such that $i = j$ since \rn{Clash} does not apply
are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Decomp} does not apply.
Thus, 
if $\const{C}_i( t_1, \ldots, t_n ) \in \ec{y}$,
$\Val$ was constructed by applying a sequence of substitutions
that replace $v_{\ec{y}}$ by $\mu v_{\ec{y}}. \const{C}_i( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} )$.
Now, say that $\const{C}( t_1, \ldots, t_n )$ occurs in some equivalence class $\ec{x}$.
Then $\Val \ec{x}$ must be of the form $\mu v_{\ec{x}}. \const{C}_i( w_1, \ldots, w_n )$,
where for each $i=1,\ldots,n$, $w_i$ was obtained by applying a sequence of substitutions
of the form mentioned above to $v_{\ec{t_i}}$.
%Each $w_i$ is either $v_{\ec{x}}$,
%a variable $v_{\ec{y}}$ where $\ec{y}$ does not contain a constructor,
%or a term of the form $\mu v_{\ec{y}}. \const{C}_j( \vec u )$.
While interpreting $w_i$ in $\Val \ec{x}$,
all free occurrences of $v_{\ec{x}}$ in $\Val \ec{x} \interp{w_i}$ are replaced by $\Val \ec{x}$. \rem{better formalize.}
Thus, since $\Val \ec{x} \interp{w_i}$ and $\Val \ec{t_i}$ are obtained by applying the same sequence of substitutions to $v_{\ec{t_i}}$,
we have $\Val \ec{w_i} = \Val \ec{x} \interp{ w_i }$.
Since $\Val'$ is obtained by applying the same substitution to all terms in the range of $\Val$,
we have that $\Val' \ec{x} \interp{w_i} = \Val' \ec{t_i}$,
and hence $\Val^\ast \ec{x} \interp{w_i} = \Val^\ast \ec{t_i}$.

\rem{TODO}
\qed
\end{proof}

>>>>

In this update, we choose an unassigned variable in $\Val$
and assign it a value that is not bisimilar to one occurring in $\mathcal{T}^\M_x(\Val)$.
Since this update removes one variable from the set $FV( \Val )$ and does not add any additional variables to $FV( \Val )$, 
it can only be applied a finite number of times.
We then let $\Val^\ast \ec{x}$ be $\nf{(\Val' \ec{x})}$ for all $\ec{x}$.

\begin{example}
Let $\tau$ be a type in $\Codata$ with three constructors $\const{A}$, $\const{B}$ and $\const{C}$ each having type $\tau \rightarrow \tau$.
Let $\Ec$ be the set consisting of equalities
$\{$
$y_1 \teq \const{A}( y_2 )$,
$y_2 \teq \const{B}( x )$,
$z_1 \teq \const{C}( z_2 )$,
$z_2 \teq \const{A}( z_1 )$,
$w_1 \teq \const{A}( x )$,
$w_2 \teq \const{C}( w_3 )$,
$x \not\teq y_2$,
$\}$.
After running the calculus to saturation on $\Ec$, the mapping $\Val$ contains:
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{2.0em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
\Val \ec{y_1} & = & \mu v_{\ec{y_1}}. & \const{A}( & \mu v_{\ec{y_2}}. & \const{B}( v_{\ec{x}} ) ) &
\Val \ec{w_1} & = & \mu v_{\ec{w_1}}. & \const{A}( v_{\ec{x}} )      & \\ 
\Val \ec{y_2} & = & \mu v_{\ec{y_2}}. & \multicolumn{2}{l}{\const{B}( v_{\ec{x}} )}      & &
\Val \ec{w_2} & = & \mu v_{\ec{w_2}}. & \const{C}( v_{\ec{w_3}} ) \\
\Val \ec{z_1} & = & \mu v_{\ec{z_1}}. & \const{C}( & \mu v_{\ec{z_2}}. & \const{A}( v_{\ec{z_1}} ) ) &
\Val \ec{w_3} & = & v_{\ec{w_3}} \\
\Val \ec{z_2} & = & \mu v_{\ec{z_2}}. & \const{A}( & \mu v_{\ec{z_1}}. & \const{C}( v_{\ec{z_2}} ) ) & 
\Val \ec{x} & = & v_{\ec{x}} \\
\end{array}\]

To construct a completion of $\Val$, we must choose values for the free variables of $\Val$, $v_{\ec{x}}$ and $v_{\ec{w_3}}$.
The content of set $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$ is $\alpha$-equivalent to
\[\begin{array}{c}
\{
\mu x. \const{A}( \mu y. \const{B}( x )),
\mu x. \const{A}( \mu y. \const{B}( y )),
\mu x. \const{B}( x ),
\mu x. \const{C}( \mu y. \const{A}( x )),
\mu x. \const{A}( \mu y. \const{C}( x )),
\mu x. \const{A}( x )
\}
\end{array}\]
Say that $\M$ interprets variables in $\Ec$ based on the mapping $\Val$,
for instance, $x$ is interpretted as $\Val \ec{x}$, $y_1$ as $\Val \ec{y_1}$ and so on.
Assigning a value for $v_{\ec{x}}$ that is $\alpha$-equivalent to a value in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$
may cause values in the range of $\Val$ to become $\alpha$-equivalent, 
which in turn may cause $\M$ to be inconsistent.
For instance, say we assign $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$ for $v_{\ec{x}}$.
After this substitution, $\Val \ec{y_2}$ is $\mu v_{\ec{y_2}}. \const{B}( \mu v_{\ec{x}}. \const{B}( v_{\ec{x}} ) )$, 
which has normal form $\mu v_{\ec{y_2}}. \const{B}( v_{\ec{y_2}} )$,
which is $\alpha$-equivalent to $\mu v_{\ec{x}}. \const{B}( v_{\ec{x}} )$.
However, this contradicts the disequality $x \not\teq y_2$ in $\Ec$.

Inversely, if the value assigned to $v_{\ec{x}}$ is not $\alpha$-equivalent to any term in $\mathcal{T}^\M_{v_{\ec{x}}}( \Val )$,
then the values in the range of $\Val$ remain distinct.
For instance, we may assign values such as 
$\mu v_{\ec{x}}. \const{C}( v_{\ec{x}} )$,
$\mu v_{\ec{x}}. \const{C}( \mu x_1. \const{A}( x_1) )$, and
$\mu v_{\ec{x}}. \const{B}( \mu x_1. \const{A}( v_{\ec{x}} ) )$
to $v_{\ec{x}}$.
Notice that legal substitutions for $v_{\ec{x}}$ may cause the range of $\Val$ to contain non-normal terms,
as in after applying the third substitution mentioned above,
$\Val \ec{w_1}$ is $\mu v_{\ec{w_1}}. \const{A}( \mu v_{\ec{x}}. \const{B}( \mu x_1. \const{A}( v_{\ec{x}} ) ) )$.
$\Box$
\end{example}

We first state the following properties of $\Val^\ast$,.
%where we write $\muvar( t )$ to denote $t$ if $t$ is a variable, or $x$ if $t$ is of the form $\mu x. u$.

\begin{lemma} 
\label{lem:model-completion}
If $\Val$ is constructed for a saturated $\Ec$,
and $\Val^\ast$ is a completion of $\Val$ for normal model $\M$, then:
\begin{enumerate}
\item $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$, for all $\ec{x}$ of type $\tau$ of $\Ec$,
\item
If $\const{C}( t_1, \ldots, t_n ) \in \ec{x}$,
then $\Val^\ast \ec{x} = \mu v_{\ec{x}}. \const{C}( w_1, \ldots, w_n )$
and $\Val^\ast \ec{x} \interp{ w_i } = \Val^\ast \ec{t_i}$ for $i = 1, \ldots, n$.
%and $\muvar( w_i ) = v_{\ec{t_i}}$ for $i = 1, \ldots, n$,
%\item
%$t_1 \interp{ t_2 } = \Val^\ast \ec{x}$
%for all $t_1$ with subterm $t_2$ in the range of $\Val^\ast$ where $\muvar(t_2)=v_{\ec{x}}$,
\item 
$\Val^\ast \ec{x} \vsim \Val^\ast \ec{y}$ if and only if $\ec{x} = \ec{y}$.
\end{enumerate}
\end{lemma}
\begin{proof}
To show 1, we first show that $\Val^\ast$ contains no free variables.
Assume by contradiction that $\Val^\ast$ contained a free variable $v_{\ec{y}}$ for some $\ec{y}$ of type $\tau$.
Then it must be the case that $\ec{y}$ does not contain a constructor term,
or else $v_{\ec{y}}$ would not occur as a free variable in $\Val$.
Consider the case when $\tau$ is finite.
By assumption, $\tau \not\in \Nondata$.
Since \rn{Split} does not apply to $\Ec$, we have $\tau \not\in \Data$.
If $\tau \in \Codata$, then by Lemma~\ref{lem:corecursive-singletons},
the cardinality of $\tau$ must be one.
Since \rn{Singleton} does not apply to $\Ec$,
there is only one equivalence class of type $\tau$ in $\Ec$,
and thus there are no terms in $\mathcal{T}^\M_x(\Val^\ast)$ of type $\tau$.
This is a contradiction, since our model completion may assign the value in the domain of $\tau$ to $v_{\ec{x}}$.
Now, consider the case when $\tau$ is infinite.
This is also a contradiction, 
since there are only a finite number of closed terms in $\mathcal{T}^\M_x(\Val^\ast)$,
and thus our model completion can assign a value not occurring in $\mathcal{T}^\M_x(\Val^\ast)$ to $v_{\ec{y}}$.
By construction, $\Val^\ast \ec{x}$ is normal.
Since \rn{Cyclic} does not apply, $\Val \ec{x}$ is acyclic when $\tau \in \Data$.
Moreover, our construction of $\Val^\ast$ applies substitutions of the form 
$\{ y \mapsto t \}$, where $t$ is acyclic when the type of $y$ is not a codatatype.
Thus, $\Val^\ast \ec{x}$ is acyclic when $\tau \not\in \Codata$.
Therefore, by definition, $\Val^\ast \ec{x}$ is bisimilar with a value in $\M( \tau )$.

To show 2, 
first note that since \rn{Clash} does not apply to $\Ec$,
for all equivalence classes $\ec{y}$,
each pair of constructor terms $\const{C}_i( \vec t )$ and $\const{C}_j( \vec u )$ in $\ec{y}$
are such that $i = j$ since \rn{Clash} does not apply to $\Ec$,
and are such that $\ec{ \vec t } = \ec{ \vec u }$ since \rn{Decomp} does not apply to $\Ec$.
Thus, 
if $\const{C}_i( t_1, \ldots, t_n ) \in \ec{y}$,
$\Val$ was constructed by applying a sequence of substitutions
that replace $v_{\ec{y}}$ by $\mu v_{\ec{y}}. \const{C}_i( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} )$.
Now, say that $\const{C}( t_1, \ldots, t_n )$ occurs in some equivalence class $\ec{x}$.
Then $\Val \ec{x}$ must be of the form $\mu v_{\ec{x}}. \const{C}_i( w_1, \ldots, w_n )$,
where for each $i=1,\ldots,n$, $w_i$ was obtained by applying a sequence of substitutions
of the form mentioned above to $v_{\ec{t_i}}$.
%Each $w_i$ is either $v_{\ec{x}}$,
%a variable $v_{\ec{y}}$ where $\ec{y}$ does not contain a constructor,
%or a term of the form $\mu v_{\ec{y}}. \const{C}_j( \vec u )$.
While interpreting $w_i$ in $\Val \ec{x}$,
all free occurrences of $v_{\ec{x}}$ in $\Val \ec{x} \interp{w_i}$ are replaced by $\Val \ec{x}$. \rem{better formalize.}
Thus, since $\Val \ec{x} \interp{w_i}$ and $\Val \ec{t_i}$ are obtained by applying the same sequence of substitutions to $v_{\ec{t_i}}$,
we have $\Val \ec{w_i} = \Val \ec{x} \interp{ w_i }$.
Since $\Val'$ is obtained by applying the same substitution to all terms in the range of $\Val$,
we have that $\Val' \ec{x} \interp{w_i} = \Val' \ec{t_i}$,
and hence $\Val^\ast \ec{x} \interp{w_i} = \Val^\ast \ec{t_i}$.

To show 3,
first note that since \rn{Bisimilar} does not apply,
$\Val \ec{x} \sim \Val \ec{y}$ if and only if $\ec{x} = \ec{y}$.
Since \rem{TODO}
we have $\nf{(\Val \ec{x})} \sim \nf{(\Val \ec{y})}$ if and only if $\ec{x} = \ec{y}$.
We show that for any substitution $\sigma$ applied to $\Val$ while constructing its completion is such that
$\nf{(\Val \ec{x} \sigma)} \sim \nf{(\Val \ec{y} \sigma)}$ if and only if $\ec{x} = \ec{y}$.
\rem{TODO}
\qed
\end{proof}
